---
title: "Derived variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Derived variables}     
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

recodeflow supports derived variables. This can be any custom function as long as the variable can be calculated on a per row bases. Currently functions requiring operations on the full data set are not supported. Two most common uses for derived variables are: 

- Variable mapping - mapping two or more variables into a single variable.
- Variables that derived using math equations - a simple example from survival pbc data is calculating copper*chol. 

`recodeflow` is capable of calculating these more complex derived variables using functions that are referenced in `variable_details.csv` within `RecTo` section with the prefix 'Func::'. The variables used in the function are referenced in the `variableStart` section with the prefix 'DerivedVar::'. For example, (`example_der`) includes `Func::example_der_fun` in the `RecTo` section; and `DerivedVar::[chol, copper]` in the `variableStart` section, which indicates the two starting variables (`chol, copper`).

## REMOVED bmi example due to it being too specific



In order to derive variables, you must load the existing custom function associated with the derived variable.
Below example_der_fun is shown. NOTE: the function name must be identical to the function name in variables and variable details, and the order and the number of passed arguments must match. 
```{r, warning=FALSE, message=FALSE}
#example_der_fun caluclates chol*copper
#@param chol the row value for chol
#@param copper the row value for copper
example_der_fun <- function(chol, copper){
  # as numeric is used to coerce in case categorical numeric variables are used.
  # Warning either chol or copper being NA will ressult in NA return
  example_der <- as.numeric(chol)*as.numeric(copper)
  
  return(example_der)
}
```


Final warning the variable used to create the derived variable must also be present in the variables and variable details.

To use the custom function they must be loaded in your environment which can be done by simply entering your functions into the console to load them, or attaching them to your own package and loading that package using "library("name of package")" or by using the rec_with_table parameter to pass the path to your function R script.
