---
title: "Derived variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Derived variables}     
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

_recodeflow_ supports the use of derived variables. This can be any custom function as long as the variable can be calculated on a per row bases. Functions requiring operations across rows or on the full data set are not supported. 

Two most common uses for derived variables are: 

- Variable mapping - mapping two or more variables into a single variable.
- Variables that derived using math equations (e.g., BMI is calculated by dividing weight by the square of height). 

To create derived variables, you need to complete two steps.

1. Create and load a customized function.
2. Add the derived variable to the var_details and var_sheet.

## Example of a derived function

We'll walk through an example of creating a derived variable with our example data.

Our customized derived function is multiplying the blood concentration of cholesterol (`chol`) with the blood concentration of bilirunbin (`bili`).

###1. Create and load a customized function for your derived variables.

Here is the customized function for our derived variable (`chol`*`bili`):
```{r, warning=FALSE, message=FALSE}
#example_der_fun caluclates chol*bili
#@param chol the row value for chol
#@param bili the row value for bili
#@export 
example_der_fun <- function(chol, bili){
  # as numeric is used to coerce in case categorical numeric variables are used.
  # Warning either chol or bili being NA will result in NA return
  example_der <- as.numeric(chol)*as.numeric(bili)
  
  return(example_der)
}
```

**Note:** You **must** use roxygen2 documentation for custom functions otherwise the function can not be attached to a package. See [roxygen2](https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html) on how to format and document your function.


Now load the customized function to your R environment. If you don't load the customized function you cannot create the derived variable. 

Load the customized function by either:

- entering your functions into the console and running the code, or 
- attaching the functions to your own package using the build and install tool and loading that package using "library("name of package")" or by using the `rec_with_table` parameter to pass the path to your function R script.


###2. Add the derived variable to the var_details and var_sheet.

Add the derived variable to the `var_sheet`. You'll use the same nomenclature as any other variable ([`variables_sheet`](../reference/variable_sheet.html)).

Add the derived variable to the variable_details. See the article [`variable_details`](../reference/variable_sheet.html) for nomenclature. 




