---
title: "Recode variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to guides}     
  %\VignetteEngine{knitr::rmarkdown}  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
source("../R/test_data_generator.R")
```

We have examples to demonstrate how to recode variables with the _recodeflow_ function [`rec_with_table()`](../reference/rec_with_table.html)

**Our examples use following packages.**

Package _recodeflow_ 

Steps on how to install `recodeflow` are in [how to install](../how_to/install.html)
```{r results= 'hide', message = FALSE, warning=FALSE}
#Load the package
library(recodeflow)
```

Package [dplyr](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8) to combine datasets (function: `bind_rows`).

```{r results = 'hide', message = FALSE, warning =FALSE}
library(dplyr)
```

**Our examples use example data**

Our examples use the dataset `pbc` from the package [survival](https://cran.r-project.org/web/packages/survival/index.html). We've split this dataset into two (tester1 and tester2) to mimic real data e.g., the same survey preformed in separate years. We've also added columns (`agegrp5` and `agegrp10`) to this dataset for our examples.


## Example 1. Recode a single variable from a single dataset

In our example datasets, the variable `sex` contains the values: m for males and f for females. 

Using dataset tester1, we'll recode the variable `sex` into a harmonized `sex` variable. The harmonized `sex` variable has the values: 0 for males and 1 for females. 

1) Recode the `sex` variable in tester1.
```{r}
sex_1 <- rec_with_table(data = tester1, 
                        variables = "sex", 
                        variable_details = variable_details,
                        log = TRUE,
                        var_labels = c(sex = "sex")
                        )
```

```{r, echo=FALSE}
head(sex_1)
```

## Example 2. Recode a single variable across multiple datasets

We'll recode and combine the variable `sex` for our two datasets. 

1) Recode the `sex` variable in tester1 and tester2.
```{r, warning = FALSE}
sex_1 <- rec_with_table(data = tester1, 
                        variables = "sex", 
                        variable_details = variable_details, 
                        log = TRUE,
                        var_labels = c(sex = "Sex")
                        )
head(sex_1)

sex_2 <- rec_with_table(data = tester2, 
                        variables = "sex", 
                        variable_details = variable_details, 
                        log = TRUE,
                        var_labels = c(sex = "Sex")
                        )
tail(sex_2)
```

2) Combine the harmonized sex variable from tester1 to the harmonized sex variable in tester2.
```{r, warning=FALSE}
sex_combined <- bind_rows(sex_1, sex_2)
```

```{r, echo=FALSE}
head(sex_combined)
tail(sex_combined)
```

3) Set labels

Labels are lost during the database merging. 

Use `set_data_labels()` to label the variables in your final dataset. `set_data_labels()` sets the labels with the original information in `variables` and `variable_details`.

```{r}
labeled_sex_combined <- set_data_labels(
  data_to_label = sex_combined,
  variable_details = variable_details,
  variables_sheet = variables
)
```

## Example 3. Recode a single variable, with different categories, from multiple datasets {#example3}

You could have a situation where a variable is the same across datasets but its categories change. 

In our example data the variable `agegrp` is different in tester1 and tester2. 

- In tester1 the `agegrp` variable is 5-year age groups: 20-24, 25-29, 30-34, etc.
- In tester2 the `agegrp` variable is 10-year age groups: 20-29, 30-39, 40-49, etc.

There are three options to facilitate the use of variables with inconsistent categories across datasets. 

### Option 1: recode category `agegrp` variable into a common variable for only datasets with the same category responses

Recode the `agegrp` variable into a common variable only in datasets were the categories are the same. If the categories are different between datasets, separate columns will be created. 

The categories in the `agegrp` variable in tester1 are different than the categories of `agegrp` in tester2. Therefore, it is not possible to have the same `agegrp` categories across our example data sets.

1) Recode `agegrp5` in tester1 and recode `agegrp10` in tester2.
```{r, warning=FALSE}
agegrp_1 <- rec_with_table(data = tester1, 
                           variables = "agegrp5", 
                           variable_details = variable_details, 
                           log = TRUE, 
                           )
head(agegrp_1)


agegrp_2 <- rec_with_table(data = tester2, 
                             variables = "agegrp10", 
                             variable_details = variable_details, 
                             log = TRUE)
head(agegrp_2)

```

2) Combine the harmonized variable `agegrp5` in tester1 with the harmonized `agegrp10` in tester2.
```{r, warning=FALSE}
agegrp_combined <- bind_rows(agegrp_1, agegrp_2)
```

```{r, echo=FALSE}
head(agegrp_combined)
tail(agegrp_combined)
```


### Option 2: recode the categorical `agegrp` variable into a continuous `age_cont` variable

Recode categorical variable `agegrp` into a single harmonized continuous
variable `age_cont`. 

`age_cont` takes the midpoint age of each category for 'agegrp' across datasets. With this option, the categorical variable 'agegrp' from each dataset can be combined into a single dataset.

1) Recode variable `agegrp` in tester1 and `agegrp` in tester2 to the harmonized continuous variable `age_cont`. 
```{r, warning=FALSE}
agegrp_1_cont <- rec_with_table(data = tester1, 
                             variables = "age_cont", 
                             variable_details = variable_details, 
                             log = TRUE)
head(agegrp_1_cont)


agegrp_2_cont <- rec_with_table(data = tester2, 
                             variables = "age_cont", 
                             variable_details = variable_details, 
                             log = TRUE)
head(agegrp_2_cont)

```

2) Combine the harmonized continous variable `age_cont` from tester1 and tester2.
```{r, warning= FALSE}
agegrp_cont_combined <- bind_rows(agegrp_1_cont, agegrp_2_cont)
```

```{r, echo=FALSE}
head(agegrp_cont_combined)
tail(agegrp_cont_combined)
```


### Option 3: recode the categorical `agegrp` variable into a harmonized categorical variable

Dataset tester1 has 5-year age groups (e.g., 30-34, 35-39), and tester2 has 10-year age groups (e.g., 30-39). Therefore, we can collapse the 5-year age groups in dataset tester1 to the same 10-year age groups in dataset tester2.

1) Recode variable `agegrp` in tester1 into `agegrp10`. recode variable `agegrp` in tester2 into `agegrp10`.
```{r, warning=FALSE}
agegrp10_1 <- rec_with_table(data = tester1, 
                             variables = "agegrp10", 
                             variable_details = variable_details, 
                             log = TRUE)
head(agegrp10_1)


agegrp10_2 <- rec_with_table(data = tester2, 
                             variables = "agegrp10", 
                             variable_details = variable_details, 
                             log = TRUE)
head(agegrp10_2)

```

2) Combine the harmonized categorical variable `age_cat` from tester1 and tester2.
```{r, warning= FALSE}
agegrp10_combined <- bind_rows(agegrp10_1, agegrp10_2)
```

```{r, echo=FALSE}
head(agegrp10_combined)
tail(agegrp10_combined)
```


## Example 4. Recode multiple variables from multiple datasets {#example4}

The variables argument in `rec_with_table()` allows multiple variables to be
recoded from a dataset. 

In this example, the `age` and `sex` variables from the tester1 and tester2 datasets will be recoded and labeled using `rec_with_table()`. 

We'll then combine the two recoded datasets into a single dataset and labeled
using `set_data_labels()`.

1) Recode `age` and `sex` in dataset tester1 and tester2
```{r, warning=FALSE}
age_sex_1 <- rec_with_table(data = tester1, 
                            variables = c("age", "sex"), 
                            variable_details = variable_details,
                            log = TRUE, 
                            var_labels = c(age = "Age", sex = "Sex")
                            )
head(age_sex_1)

age_sex_2 <- rec_with_table(data = tester2, 
                            variables = c("age", "sex"), 
                            variable_details = variable_details,
                            log = TRUE, 
                            var_labels = c(age = "Age", sex = "Sex")
                            )
head(age_sex_2)
```

2) Combine the harmonized variables `age` and `sex` from tester1 and tester2.
```{r, warning=FALSE}
combined_age_sex <- bind_rows(age_sex_1, age_sex_2)
head(combined_age_sex)
```

3) Set labels

Use `set_data_labels()` to label the variables in your final dataset. `set_data_labels()` sets the labels with the original information in `variables` and `variable_details`.

`var_labels` can be used all the variables in `variables.csv` or a subset of variables.

```{r, warning=FALSE}
labeled_combined_age_sex <- 
  set_data_labels(
      data_to_label = combined_age_sex,
      variable_details = variable_details, 
      variables_sheet = variables
      )
```

You can check if labels have been added to your recodeed dataset by using `get_label()`.

```{r, warning=FALSE}
library(sjlabelled) 
get_label(labeled_combined_age_sex)
```

For more information on `get_label()` and other label helper functions, please
refer to the [sjlabelled](https://strengejacke.github.io/sjlabelled/reference/)
package.


## Example 5. Recode all variables in the variables worksheet {#example5}

All the variables listed in `variables` worksheet can be recoded with `rec_with_table()`. 

In this example, all variables specified in the `variables` worksheet will be recoded and combined for the datasets tester1 and tester2.

```{r, echo=FALSE}
options(htmlwidgets.TOJSON_ARGS = list(na = "string"))
```

1) Recode all variables listed in the variables worksheet, for dataset tester1 and dataset tester2
```{r, warning=FALSE}
recoded1 <- rec_with_table(data = tester1,
                          variables = variables,
                          variable_details = variable_details,
                          log = TRUE,
                          )

recoded2 <- rec_with_table(data = tester2,
                          variables = variables,
                          variable_details = variable_details,
                          log = TRUE,
                          )
```

2) Combine recoded datasets
```{r, warning=FALSE}
combined_dataset <- bind_rows(recoded1, recoded2)
```

3) Set labels for the combined recoded dataset
```{r}
labeled_combined <- set_data_labels(data_to_label = combined_dataset,
                                    variable_details = variable_details,
                                    variables_sheet = variables
                                    )
```

## Example 6: Add the data origin in combined datasets {#example6}

To know the origin of each row of data, you can use the `rec_with_table` argument 
`attach_data_name`. When the argument `attach_data_name` is set to true it will add a column with the name of the dataset the row is from.

1) Recode variables `age` and `sex` and attach dataset name for tester1 and tester2.
```{r, warning=FALSE}
age_sex_1 <- rec_with_table(data = tester1,
                            variables = c("age", "sex"), 
                            variable_details = variable_details,
                            var_labels = c(age = "Age", sex = "Sex"),
                            log = TRUE,
                            attach_data_name = TRUE
                            )

age_sex_2 <- rec_with_table(data = tester2,
                            variables = c("age", "sex"), 
                            variable_details = variable_details,
                            var_labels = c(age = "Age", sex = "Sex"),
                            log = TRUE,
                            attach_data_name = TRUE
                            )
```

2) Combine the harmonized datasets
```{r, warning=FALSE}
combined_age_sex <- bind_rows(age_sex_1, age_sex_2)

head(combined_age_sex)
tail(combined_age_sex)
```


## Example 7. Recode derived variables {#example7}

Derived variables are variables that are not in the original dataset; rather they are created using variables the original dataset.

Descriptions of derived functions are in the article [derived functions](../articles/derived_variables.html)


To recode a derived variable, you must: 

- create a customized function,
- defined the derived variable on the worksheets `variables` and `variable_details`,
- recode the variables that make up the derived variable. 

Our example derived variable `example_der` equals `chol` times `bili`.

1) Recode the underlying variables: `chol` and `bili` and the derived variable `example_der` for tester1 and tester2.
```{r, warning=FALSE}
derived1 <- rec_with_table(data = tester1,
                          variables = c("chol", "bili","example_der"),
                          variable_details = variable_details,
                          log = TRUE)

derived2 <- rec_with_table(data = tester2,
                          variables = c("chol", "bili","example_der"),
                          variable_details = variable_details,
                          log = TRUE)
```

2) Combine the harmonized variables: `chol`, `bili`, and `exampler_der`
```{r, warning=FALSE}
combined_der <- bind_rows(derived1, derived2)
```
