---
title: "How to use recodeflow with your data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to add variables to cchsflow}     
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

source("../R/test_data_generator.R")
```


# How to use `recodeflow` with your data

To use `recodeflow` with your data, you'll need create the following worksheets:

- `variable_details` - mapping of variables across datasets and a list of instructions on how to transform variables
- `variable_sheet` - a list of variables to transform, and

Files can either be a `.csv` file that you import to R or create directly in R as a dataframe.

If you have derived variables, you will also need to create custom functions. For more information on custom functions see the article [derived_variables](../reference/derived_variables.html).


## Example data

Our examples use the dataset `pbc` from the package [survival](https://cran.r-project.org/web/packages/survival/index.html). We've split this dataset into two (tester1 and tester2) to mimic real data e.g., the same survey preformed in separate years. We've also added columns (`agegrp5` and `agegrp10`) to this dataset for our examples.

We'll use our example datasets and the variable `stage`.
The variable `stage` captures the histologic stage of the patient's disease (primary biliary cirrhosis (PBC)).

## How to create the `variable_details` worksheet

The `variable_details` worksheet does two important steps. First, it maps variables across datasets. Second, it gives instructions on how to transform the variables. 

Note: additional information for the `variable_details` worksheet is in the article [variable_details](../reference/variable_details.html).

Note: additional details on how to add derived variables to the `var_detials` worksheet is in the article [derived_variables](../reference/derived_variables.html).


### Rows

For the `stage` variable, there are the following six rows:

- 4 rows for the 4 categories for the histological stage of the disease, 
- 1 row for the not applicable responses, and 
- 1 row for else. 

Since `stage` is coded consistently across the two datasets (tester1 and tester2) we only need one row per categorical response.

### Columns (14)

1. **variable:** the name of the final transformed variable. We'll use the same name as in the original dataset `stage` though you could rename the variable.

Write `stage` in the column **variable** in the six rows.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
kable(variable_details[58:63, 1], col.names = 'variable')
```

2. **toType:** `stage` captures the stage of the disease and is originally a categorical variable. Therefore, it does not make sense to change `stage` into a continuous variable. 

Write 'cat' in the six rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:2])
```

3. **fromType:** `stage` is originally a categorical variable.

Write 'cat' in the six rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:3])
```

4. **databaseStart:** `stage` was captured identically for both datasets. Therefore, we will transform the variable the same way regardless of which dataset it's from.

Write the dataset names, separated by a comma, in the six rows

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:4])
```

5. **variableStart:** The original variable `stage` is identical across our datasets. 

Write the variable name in squared brackets once per row, for all six rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:5])
```

6. **variableStartLabel:** 

Write "stage" in the 6 rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:6])
```

7. **numValidCat:** The number of categories within the variable `stage`. There are four categories in `stage`: 1, 2, 3, and 4. Note that for categories: not applicable, missing, and else, are not included in the category count.

Write 4 in each of the six rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:7])
```

8. **recTo:** Since `stage` is a categorical variable, you'll capture the category you are transforming each row too. For the not applicable rows `NA::a` is written. For the missing and else rows `NA::b` is written. The `haven` package is used for tagging NA in numeric variables.

We are not changing the categories. Therefore, the four stages captured in the original data. For the not applicable rows write `NA::a`. For the else rows write `NA::b`.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:8])
```

9. **catLabel:** Name the categorical level for the transformed variable.

Write Stage 1, Stage 2, Stage 3, Stage 4, NA, and missing.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:9])
```

10. **catLabelLong:** You can elaborate on the name of the categorical level of the transformed variable.

Copy values from **catLabel**

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:10])
```

11. **units:** The histologic stage of disease does not have units of measurement. 

Write "N/A" in all six rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:11])
```

12. **recFrom:** Since we are not combining levels of categories and we are keeping the category levels the same, the recFrom column will be identical to recTo. 

Write the category level you are transforming each row too. For the not applicable rows `NA::a` is written. For the missing and else rows `NA::b` is written.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:12])
```

13. **catStartLabel:** The `stage` label should be identical to what is shown in the original data documentation. For the missing rows, each missing category is described along with their coded values.

Write Stage 1, Stage 2, Stage 3, Stage 4, NA, and missing.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:13])
```

14. **notes: ** Capture any important differences in a variable across datasets. For our example, there are no differences across datasets.

Write "This is sample survival pbc data" in all six rows.

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[58:63, 1:14])
```

## `variable_details` for dervived variables

The same naming convention applies to derived variables with the exception of two columns:

1. In **variableStart**, instead of database names being listed, **DerivedVar::** is written followed with the list of original variables used inside square brackets. 
+ `DerivedVar::[var1, var2, var3]`

2. In **recTo**, write **Func::** followed with the exact name of the custom function used to create the derived variable.
+ `Func::derivedFunction`

A derived variable looks like this in `variable_details.csv`

```{r, echo=FALSE, warning=FALSE}
kable(variable_details[64,1:14])
```

## How to create the variables worksheet `variables_sheet`

Once mapped and specified on `variable_details`, the `stage` variable can be specified on the variables worksheet: `variables_sheet`. Ensure that the names you used in the `variable_details` worksheet are identical to those listed in `variables_sheet`.

1. **variable:** the name of the transformed variable (variable_details worksheet column: 'variable')

2. **label:** the shorthand label for the variable (variable_details worksheet column: '*variableStartShortLabel*')

3. **labelLong:** a more detailed label for the variable (new column, it is not in the variable_details worksheet)

4. **section:** group of data the variable belongs to (new column, it is not in the variable_details worksheet)

5. **subject:** sub-group of data the variable belongs to (new column, it is not in the variable_details worksheet)

6. **variableType:** whether the transformed variable is categorical or continuous (variable_detail worksheet column: 'toType')

7. **units:** any units for the variable (variable_detail worksheet column 'units')

8. **databaseStart:** the list of databases that contain the variable of interest (variable_detail worksheet column: 'databaseStart')

9. **variableStart** the original variable name (variable_detail worksheet column: 'variableStart')

```{r, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
kable(variables_sheet[19, ])
```


