---
title: "rec_with_table"
output:
  html_document:
    df_print: paged
vignette: |
  %\VignetteIndexEntry{rec_with_table}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
source("../R/test_data_generator.R")
```


# What is the function `rec_with_table`?

`rec_with_table` is the function used to recode variables within your dataset into harmonized variables. `rec_with_table` reads:

- `variable` for a list of variables to recode, and
- `variable_details` for instructions on how to recode the variable(s)

```{r}
library(recodeflow)
status <- rec_with_table(data = tester1, 
                         variables = "status", 
                         variable_details = variable_details,
                         log = TRUE,
                         var_labels = c(status = "Status"))
```

# Output - Warning messages

Warning messages will appear when your dataset is missing variables in your
`variable` worksheet or `rec_with_table()` call.

# Output - Notes

By default, `rec_with_table()` prints notes to the console.

Notes can give important details on what and how variables are harmonized. For example, notes can include details on:

- which datasets include a variable,
- how a variable changes across datasets, and 
- if the recoding rules for a variable is dependent on which dataset the variable is in

You should read and consider `notes` before deciding whether to harmonize a variable.

The notes output is from the column *notes* in the`variable_details` 

# Arguments - Required

- `data` A dataframe(s) containing data to be recoded. 

- `variables` Either a string of variables to recode or a dataframe containing the variables (e.g., `variables`)

- `variable_details` A dataframe containing the specifications for recoding. 

```{r, eval=FALSE}
#To recode only a few variables
age_sex_1 <- rec_with_table(data = tester1, 
                         variables = c("age","sex"),
                         variable_details = variable_details)


#To recode all variables listed on the variable worksheet
recoded1 <- rec_with_table(data = tester1,
                          variables = variables,
                          variable_details = variable_details)
```

# Arguments - Default 

With `rec_with_table()` the following arguments are only called if you want to modify them.

- `database_name` defaults to NULL. As a result, `rec_with_table()` uses the
  database indicated in the `data` argument.
  
- `else_value` defaults to NA. Values out of range set to NA.

- `append_to_data` defaults to FALSE. Recoded variables will not be appended
  to the original dataset.
  
- `log` defaults to FALSE. Logs of the variable recoding will not be
  displayed.
  
- `notes` defaults to TRUE. Information in the `Notes` column from
  `variable_details` worksheet will be displayed
  
- `var_labels` is set to NULL. This argument can be used to add labels to a
  recoded dataset that only contains a subset of variables from
  `variables` worksheet. The format to add labels to a subset of variables is
  `c(variable = "Label")`.
  
- `attach_data_name` is set to FALSE. This argument can be used to append a
  column that denotes the name of the dataset each recoded row came from.



