---
title: Derived Variables Scope 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = TRUE,
  message = TRUE,
  error = FALSE
)

# Load required packages
library(haven)
library(reticulate)
library(knitr)
library(dpylr)
```

## Objective

The objective of this document is to define the scope and specifications for
derived variables in the library.

## Context

The variable details sheet within the recodeflow library enables a user to
encode the rules to create a new variable from one or more starting variables.
Broadly, any rule that maps the values of one variable into another can be
encoded in the variable details sheet. Examples include rules that map one or
more categories of a variable into a new category or a rule that maps an
interval of a continuous variable into a category. Variables that cannot be
created with simple mapping rules can be recoded by using a function. Examples
include variables that require mathematical operations or conditional logic.

Any variable that requires a function for its creation is considered a derived
variable. A good example of this is BMI which requires mathematical operations
for its calculation and optionally conditional logic to validate its inputs.
This document will go over the scope and specifications for such derived
variables and the functions used to derive them.

## Scope

### Function Language

The library should support only functions written in R. The R ecosystem is
moving towards allowing its users to execute code from other statistical
languages. An example is the
[reticulate](https://rstudio.github.io/reticulate/index.html) package which
enables the execution of arbitrary Python code within a .R file. Example below,

```{r, python-in-R}
# Python code that creates a function to calculate BMI
BMI_python_code <- "def BMI(weight, height): 
  return weight/(height*height)
" 
# Run reticulate to bring the Python function over to the R environment enabling
# us to call it using R code
reticulate_output <- reticulate::py_run_string(BMI_python_code)
# Calculate BMI using the created Python function
BMI_python <- reticulate_output$BMI(25, 5)
# Expect the calculated BMI to be 1
stopifnot(BMI_python == 1)
print(paste("Calculated BMI:", BMI_python))
```

To reduce the library scope, currently only R functions will be allowed.

### Function Inputs and Outputs

The function inputs and output should be standardized to accept either scalar or
vector values.

For example, a function that calculates BMI will take as input height and
weight. The library should make a decision on whether the inputs can be scalar
values for example 80 and 180 or vector values for example c(80) and c(180).

This decision will have implications on the performance, user experience, and
portability of the function as described below.

R being a statistical language is optimized to run on vectors. However, a
function accepting scalar inputs can be easily vectorized using the
[Vectorize](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Vectorize.html)
function available in base R.

The code below displays a table that compares the performance of calculating BMI
using:

1.  Scalar inputs
2.  Vector inputs
3.  A vectorized version of 1 created using the `Vectorize` function

```{r}
scalar_BMI <- function(weight, height) {
  bmi <- NA
  if(is.na(weight) | is.na(height)) {
    bmi <- NA
  }
  bmi <- weight/(height*height)
}

vector_BMI <- function(weight, height) {
    bmi <- ifelse(
      is.na(weight) | is.na(height),
      NA,
      weight/(height*height)
    )
}

vectorized_BMI <- Vectorize(scalar_BMI)

# Create larger test dataset for better comparison
test_data <- data.frame(
  weight = sample(1:100, 1000, replace = TRUE),
  height = sample(1:100, 1000, replace = TRUE)
)

# In milliseconds
get_func_perf <- function(func) {
  start_time <- Sys.time()
  func()
  end_time <- Sys.time()
  run_time <- end_time - start_time
  return(as.numeric(run_time)*1000)
}

# Original methods
scalar_run_time <- get_func_perf(function() {
  bmi <- c()
  for(i in 1:nrow(test_data)) {
    bmi <- c(bmi, scalar_BMI(test_data[i, "weight"], test_data[i, "height"]))
  }
})

vector_run_time <- get_func_perf(function() {
  vector_BMI(test_data$weight, test_data$height)
})

vectorized_run_time <- get_func_perf(function() {
  vectorized_BMI(test_data$weight, test_data$height)
})

# Mapply methods - testing all three functions
mapply_vectorized_run_time <- get_func_perf(function() {
  mapply(vectorized_BMI, test_data$weight, test_data$height)
})

mapply_vector_run_time <- get_func_perf(function() {
  mapply(vector_BMI, test_data$weight, test_data$height)
})

mapply_scalar_run_time <- get_func_perf(function() {
  mapply(scalar_BMI, test_data$weight, test_data$height)
})

# Dplyr methods - testing only vector_BMI and vectorized_BMI
library(dplyr)
dplyr_vector_run_time <- get_func_perf(function() {
  test_data %>% 
    mutate(bmi = vector_BMI(weight, height))
})

dplyr_vectorized_run_time <- get_func_perf(function() {
  test_data %>% 
    mutate(bmi = vectorized_BMI(weight, height))
})

calculate_percent_diff <- function(num1, num2) {
  return((abs(num1 - num2)/((num1 + num2)/2))*100)
}

# Enhanced results table with all valid methods
run_time_table <- data.frame(
  "Type" = c("Scalar", 
             "Vector", 
             "Vectorized", 
             "Mapply (vectorized_BMI)", 
             "Mapply (vector_BMI)",
             "Mapply (scalar_BMI)",
             "Dplyr (vector_BMI)",
             "Dplyr (vectorized_BMI)"),
  "Run Time in ms" = c(scalar_run_time, 
                      vector_run_time, 
                      vectorized_run_time, 
                      mapply_vectorized_run_time,
                      mapply_vector_run_time,
                      mapply_scalar_run_time,
                      dplyr_vector_run_time,
                      dplyr_vectorized_run_time),
  "Percent Difference" = c("Reference", 
                          calculate_percent_diff(scalar_run_time, vector_run_time),
                          calculate_percent_diff(scalar_run_time, vectorized_run_time),
                          calculate_percent_diff(scalar_run_time, mapply_vectorized_run_time),
                          calculate_percent_diff(scalar_run_time, mapply_vector_run_time),
                          calculate_percent_diff(scalar_run_time, mapply_scalar_run_time),
                          calculate_percent_diff(scalar_run_time, dplyr_vector_run_time),
                          calculate_percent_diff(scalar_run_time, dplyr_vectorized_run_time))
)

knitr::kable(run_time_table)
```

In summary, in the order from slowest to fastest we have scalar, vectorized, and
vector.

Ergonomically, users within R may be more comfortable with vectors since
datasets are usually in data.frames and individual columns can be plucked out as
vectors using the \$ operator. Users of non-statistical languages will be more
comfortable with scalars since they usually deal with scalar values.

Scalar values are more easily portable to non-statistical languages which
becomes a concern if the recoding rules need to be run for example in a web
browser where Javascript is the programming language of choice.

### Supported Operations

The library should allow the user to user to use any R code within the function.
For example, code that uses external libraries, sources other R files etc.

### Tagged NA

The need for tagged NA values has been described elsewhere. An issue with
derived functions is that users can easily return NA values that are not tagged.
The library should provide appropriate warnings to users when an NA value is not
tagged or tagged with an un-recognized value. For example,

```{r}
BMI_non_tagged_NA <- function(weight, height) {
  if(is.na(weight) | is.na(height)) {
    return(NA)
  }
  return(weight/height*height)
}

BMI_tagged_NA <- function(weight, height) {
  if(is.na(weight) | is.na(height)) {
    return(haven::tagged_na("a"))
  }
  return(weight/height*height)
}
```

The BMI_non_tagged_NA function when run by the library should warn the user that
they have returned an NA value that is not tagged. The BMI_tagged_NA function
should pass with no issues since the user is returning a tagged NA value. œ
### Best practices

Keeping in mind that users will be writing derived functions on their own, the
library should provide guidance on how to write good functions. This can include
but is not limited to:

-   Properly documenting the function behaviour, inputs, and outputs
-   Providing examples of running the function using different input types like
    scalars, vectors, data frames etc.
-   Providing test cases for the functions
-   Handling invalid input values